---
title: "Reproducible Research - Assignment 1"
author: "Suman Sengupta"
date: "21 Sep 2015"
output: html_document
---

## Loading and preprocessing the data

```{r}
if(!file.exists("./data")){dir.create("./data")}
#fileURL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
#download.file(fileURL,destfile= "./data/activity.csv")
#unzip(zipfile="./data/repdata-data-activity.zip",exdir="./data")
activity <- read.csv("./data/activity.csv")
```

##What is mean total number of steps taken per day?

```{r}
library(dplyr)
library(ggplot2)
library(lattice)

#Calculate total number of steps taken per day (ignore steps = NA)

activity1 <- activity %>% filter(is.na(steps) == FALSE )
activity1 <- activity1 %>% group_by(date) %>% summarize (total = sum(steps))

# Histogram to plot Total number of steps taken per day 
plt <- ggplot(activity1, aes(total)) + 
  geom_histogram(col= "red", fill = "green" , alpha = 0.2) + 
  labs (x = "Steps per Day", y = "Count", title="Histogram for Steps/Day")
plt

#Calculate and report the mean and median of the total number of 
#steps taken per day
meanstep <- mean(activity1$total)
medianstep <- median(activity1$total)

#Mean Step is:
meanstep
#Median Step is:
medianstep

#Plot Mean on the Histogram of Steps/Day
plt + geom_vline(xintercept = c(mean(activity1$total),median(activity1$total)), colour="blue")

maxsteps <- activity1 %>% filter (total == max(total)) 
maxsteps

```

##What is the average daily activity pattern?
```{r}
#Calculate the steps/interval

activity1 <- activity[complete.cases(activity),]
activity1 <- activity1 %>% group_by (interval) %>% summarize(total = sum(steps))

#Plot 5-min interval vs steps taken across all days
ggplot(activity1, aes(interval, total)) +geom_line(color = "blue") +
  labs(x = "5-min interval", y= "Total steps" ) +
  labs (title = "Time Series Total steps per 5-min interval")

# Determine which 5-minute interval contains the maximum number of steps
maxinterval <- activity1 %>% filter (total == max(total)) %>% select(interval)
maxinterval
```

## Inputting Missing values

```{r}
# Calculate and report the total number of missing values in the dataset
nrow(activity) - nrow(activity[complete.cases(activity),])

#Input missing values - Mean of corressponding day will be plugged in

activity1 <- activity[complete.cases(activity),]
activity1 <- activity1 %>% group_by (interval) %>% summarize(total = mean(steps))
#Append a column with the 
mergeddata <- merge(activity, activity1, by = "interval")

#Identify rows with missing values
ind <- is.na(mergeddata$steps)

filleddata <- mergeddata
filleddata[ind,2] <- filleddata[ind,4]
filleddata <- filleddata[1:3]

# Calculate total number of steps taken per day
filleddata1 <- filleddata %>% group_by(date) %>% summarize (total = sum(steps))

# Histogram to plot Total number of steps taken per day 
plt <- ggplot(filleddata1, aes(total)) + 
  geom_histogram(col= "red", fill = "green" , alpha = 0.2) + 
  labs (x = "Steps per Day", y = "Count", title="Histogram for Steps/Day")
plt


#steps taken per day
meanstep <- mean(filleddata1$total)
medianstep <- median(filleddata1$total)

#Mean Step is:
meanstep
#Median Step is:
medianstep
```
What is the impact of imputing missing data on the estimates of the total daily number of steps?

There is not much impact after injecting the missing value.

```{r}
maxsteps <- filleddata1 %>% filter (total == max(total)) 
maxsteps
```
##Are there differences in activity patterns between weekdays and weekends?

```{r}
# Determine if Date is a weekday or weekend
tempdata <- filleddata
tempdata$weekend <- ifelse(weekdays(as.Date(filleddata$date)) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
tempdata <- tempdata %>% group_by (interval,weekend) %>% summarize(total = sum(steps))
head(tempdata)

# Panel plot containing a time series plot (i.e. type = "l") of the 
#5-minute interval (x-axis) and the average number of steps taken

xyplot(total ~ interval | weekend, data = tempdata, layout = c(1, 2), type = "l")
```

